Describe and Explore: Using feature-based time series clustering to identify meaningful structures in intensive longitudinal data

### 2023-01-03 manuscript.tex
:pencil: intro explainability

\begin{mdframed}[style=mdfbox]
\noindent\center\textit{Structure}:
\begin{itemize}[nosep]
    \item introduction
    \begin{itemize}[nosep]
        \item increase in number and variety of ESM data + ts clustering to understand complex developmental differences.
        \item why cluster time series? 
        \begin{enumerate}[nosep]
            \item differences between clusters
            \item patterns within clusters
            \item relation to out-of cluster variables
            \item robust theories and interventions
        \end{enumerate}
        \item can't cluster all the raw data (b/c too much data and not directly comparable).
        \item summarize the data and cluster based on the summaries.
        \item ideally these summaries are:
        \begin{enumerate}[nosep]
            \item good at captuing the data
            \item meaningful = help us understand how the clustering arrived at the output (i.e., explainable) + helps us make sense of the final cluster output (i.e., interpretable)
        \end{enumerate}
        \item currently model parameters as summaries.
        \item limitations of parameter summaries: assumptions not always good at capturing data and model parameters not always interpretable
        \item current solution: more complicated models (e.g., address violated assumptions or add different types of parameters)
        \item proposed solution: features as flexible and meaningful approach of summarizing the time series
        \item ---------------------------------------------------------------------------
        \item limitations of model parameters: 
        \begin{itemize}[nosep]
            \item strong assumptions = not always a good (dis-)similiarity: 
            \begin{enumerate}[nosep]
                \item non-equidistant measurements
                \item non-imputable missingness
                \item non-stationary
                \item non-linear developments
            \end{enumerate}
            \item difficult to interpret psychologically 
            \begin{enumerate}[nosep]
                \item limited to few ways of psychological effects (e.g., lagged + cross-lagged effects)
            \end{enumerate}
        \end{itemize} 
        \item push for more complicated models aimed at addressing (1) similarity concerns (e.g., amend assumptions; e.g., see den Teuling) or (2) interpretatbility (e.g., more types of effects; e.g., see krone2018)
        \item proposed alternative: go back to interpretable similarity measurements. General statistics and parameters instead of only parameters.
    \end{itemize}
    \item Feature-based Time Series Clustering
    \begin{itemize}[nosep]
        \item different types of similarity grouping: five types
        \item why features: flexible similarity (b/c more variety), less assumptions, easy to calculate/extract, theory-based selection (as per RQ), psychological interpretability, ...
        \item Features in Psychological Time Series
        \begin{itemize}[nosep]
            \item central tendency
            \item variability
            \item (in)stability
            \item self-similarity
            \item linear trend
            \item nonlinearity
        \end{itemize}
        \item Analysis Steps
        \begin{itemize}[nosep]
            \item input variables
            \item feature extraction
            \item (feature reduction)
            \item feature clustering
            \item cluster evaluation
        \end{itemize}
    \end{itemize}
    \item Empirical Illustration
    \item Discussion (summary, limitation, implications, conclusion, etc.)
\end{itemize}
\vspace{1em}
\end{mdframed}

@@

However, while clustering can be incredibly useful, arriving at these clusters critically depends on two core challenges.

this means time series clustering critically depends on making time series comparable to 
identify key (dis)similarities and then accurately distinguishing different groups \citep[e.g.,][]{Aghabozorgi2015}.

The main aim of time series clustering is to inductively group participants in such a way that the dynamics of some variable(s) have maximum similarity for individuals in the same group, and minimum similarity for individuals in different groups \citep[e.g.,][]{Aghabozorgi2015}. In practice, this means time series clustering critically depends on identifying similarities or dissimilarities in time series and then accurately distinguishing different groups \citep[e.g.,][]{Aghabozorgi2015}. Broadly speaking, time series (dis-)similarity can be approximated by either comparing the raw time series or by comparing some numerical representation of the time series (e.g., \citealp[]{liao2005}; see \fgrref{fig:tsClustTax} for an illustration). The (dis-)similarity measurements can then be clustered either on their own or as part of the representation estimation (e.g., \citealp[]{denteuling2021}; see `embedded' in \fgrref{fig:tsClustTax}).

@@

>>>>

\begin{mdframed}[style=mdfbox]
\noindent\center\textit{Structure}:
\begin{itemize}[nosep]
    \item introduction
    \begin{itemize}[nosep]
        \item increase in number and variety of ESM data + ts clustering to understand complex developmental differences.
        \item why cluster time series? 
        \begin{enumerate}[nosep]
            \item differences between clusters
            \item patterns within clusters
            \item relation to out-of cluster variables
            \item robust theories and interventions
        \end{enumerate}
        \item can't cluster all the raw data (b/c too much data and not directly comparable).Instead, summarize the data and cluster based on the summaries.
        \item ideally these summaries are:
        \begin{enumerate}[nosep]
            \item good at captuing the data
            \item meaningful = help us understand how the clustering arrived at the output (i.e., explainable) + helps us make sense of the final cluster output (i.e., interpretable)
        \end{enumerate}
        \item currently model parameters as summaries.
        \item limitations of parameter summaries: assumptions not always good at capturing data and model parameters not always interpretable
        \item current solution: more complicated models (e.g., address violated assumptions or add different types of parameters)
        \item proposed solution: features as flexible and meaningful approach of summarizing the time series
        \item ---------------------------------------------------------------------------
        \item limitations of model parameters: 
        \begin{itemize}[nosep]
            \item strong assumptions = not always a good (dis-)similiarity: 
            \begin{enumerate}[nosep]
                \item non-equidistant measurements
                \item non-imputable missingness
                \item non-stationary
                \item non-linear developments
            \end{enumerate}
            \item difficult to interpret psychologically 
            \begin{enumerate}[nosep]
                \item limited to few ways of psychological effects (e.g., lagged + cross-lagged effects)
            \end{enumerate}
        \end{itemize} 
        \item push for more complicated models aimed at addressing (1) similarity concerns (e.g., amend assumptions; e.g., see den Teuling) or (2) interpretatbility (e.g., more types of effects; e.g., see krone2018)
        \item proposed alternative: go back to interpretable similarity measurements. General statistics and parameters instead of only parameters.
    \end{itemize}
    \item Feature-based Time Series Clustering
    \begin{itemize}[nosep]
        \item different types of similarity grouping: five types
        \item why features: flexible similarity (b/c more variety), less assumptions, easy to calculate/extract, theory-based selection (as per RQ), psychological interpretability, ...
        \item Features in Psychological Time Series
        \begin{itemize}[nosep]
            \item central tendency
            \item variability
            \item (in)stability
            \item self-similarity
            \item linear trend
            \item nonlinearity
        \end{itemize}
        \item Analysis Steps
        \begin{itemize}[nosep]
            \item input variables
            \item feature extraction
            \item (feature reduction)
            \item feature clustering
            \item cluster evaluation
        \end{itemize}
    \end{itemize}
    \item Empirical Illustration
    \item Discussion (summary, limitation, implications, conclusion, etc.)
\end{itemize}
\vspace{1em}
\end{mdframed}

@@

%The main aim of time series clustering is to inductively group participants in such a way that the dynamics of some variable(s) have maximum similarity for individuals in the same group, and minimum similarity for individuals in different groups \citep[e.g.,][]{Aghabozorgi2015}.

However, while clustering can be incredibly useful, arriving at these clusters critically depends on two core challenges. First, time series need to be made comparable in order to identify key (dis)similarities and second, comparable (dis)similarities need to be accurately distinguishing into different groups \citep[e.g.,][]{Aghabozorgi2015}. In practice, most psychological time series cannot be compared based on the raw data itself. This is the case because in most cases the raw time series include too many time data points for most clustering methods to handle \citep[e.g.,][]{altman2018} and, more importantly, individual time points are oftentimes not directly comparable between participants in psychological data and would lead to misspecifications \citep[e.g., ][]{faloutsos1994}. While such issues can be avoided with transformations for highly regular, controlled, and comparable time series such as EEG \citep[e.g.,][]{huang1985}, most ESM psychologists are usually not interested in directly comparing individual timepoints between participants but are interested in developmental patterns and relationships. As a result, most psychological time series are summarized via a numerical representation and these numerical summaries are then comparable and used to cluster participants (e.g., \citealp[]{liao2005}; see \fgrref{fig:tsClustTax}).

@@

The representations that summarize the original time series data should (1) capture the original data accurately without loosing too much information, and (2) should be conceptually meaningful \citep[e.g.,][]{vandermaaten2009}. Extracting accurate and meaningful representations of the time series can be essential for understanding what goes into the clustering algorithm (i.e., assists with explainability) and can be crucial in making sense of the final cluster output \citep[i.e., assists with interpretability; e.g.,][]{Kennedy2021}.
